---
title: "assignment_01"
format: html
---

# Assignment 01

* Sean Mussenden
* 2024-09-21
* Assignment 01

## Load Libraries

```{r}

# Load the tidyverse
library(tidyverse)
```

## Load Data

```{r}

# Load data provided for assignment
blackindex_master <- read_csv("https://github.com/wellsdata/CompText_Jour/raw/main/data/blackindex_master.csv")

blackindex_master


```

## Describe data

```{r}
## Instruction: Using code, describe the number of rows and columns in the dataset

print(paste0("This dataset has ", nrow(blackindex_master), " rows and ", ncol(blackindex_master), " columns. Each row in this dataset appears to describe a single article that appeared in a Black newspaper between ", min(blackindex_master$year), " and ", max(blackindex_master$year), ". The articles appear to have been sourced from several historical newspaper databases and archives. Many of the article headlines reference racial terror lynching."))

glimpse(blackindex_master)

```

## Research Questions

### Research Question/Task 01

Q: What are the top five cities in the newspaper_city column?
A:
```{r}

# Count newspapers by city
newspaper_city_count <- blackindex_master %>%
  count(newspaper_city, sort = TRUE)

# Create a df removing NA values 
newspaper_city_count_no_na <- newspaper_city_count %>%
  filter(!is.na(newspaper_city))

# Count NA values
newspaper_city_count_na <- newspaper_city_count %>%
  filter(is.na(newspaper_city)) %>%
  pull(n)

# Loop through top five cities to create a string with city and count of articles
top_five <- map(1:5, ~ paste0(newspaper_city_count_no_na$newspaper_city[.x], 
                              " (", newspaper_city_count_no_na$n[.x], ")"))

# Print out summary sentence
print(paste0("A: The top five cities in the newspaper_city column are: ", 
             paste(top_five, collapse = ", "), ". However, it's important to note that the dataset of ",nrow(blackindex_master), " rows has ", newspaper_city_count_na, " rows with a missing value in the newspaper_city column."))

# Display top five table
newspaper_city_count_no_na %>%
  head(5) %>%
  rename(count = n)


```

### Research Question/Task 02

Q: Create a table that counts all newspaper_city entries by year, filtering out the out the NA values.
A: 

```{r}

# Create required dataframe
newspaper_city_by_year <- blackindex_master %>%
  filter(!is.na(newspaper_city)) %>%
  count(newspaper_city, year, name = "count") %>%
  arrange(newspaper_city, year)

# Display dataframe
newspaper_city_by_year


```
### Research Question/Task 03

Q: Create a simple column ggplot chart that shows the total entries by year.
A: 

```{r}


# Create df grouping by year, then charting
blackindex_master %>%
  count(year) %>% 
  ggplot(aes(x = year, y = n)) +
  geom_col() +
  labs(title = "Total Articles by Year in Black Index Master",
       subtitle = "There are more articles from 1920 in this dataset than any other year",
       x = "Year",
       y = "Total Articles",
       caption = "Source: Black Index Master Dataset, Via Prof. Rob Wells") +
  scale_x_continuous(breaks = seq(1850, 2020, 10)) +
  #rotate x axis labels 45 degrees
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

## Summary

The dataset 'blackindex_master' provided by Prof. Rob Wells contains information about 1803 articles from Black newspapers. Many of the articles' headlines reference racial terror lynchings, which targeted African-Americans in the mid-to-late late 19th and early 20th centuries. The articles cover a period from 1850 to 2002, though the majority of articles are from the period between 1900 and 1950, with a clear peak in 1920. 

The top five cities for publications represented in the data are Pittsburgh, Chicago, Atlanta, Norfolk	and New York, though the very large number of NA values in the newspaper city column suggest that the column should not be used to draw conclusions about the geography of the overall dataset. Some cities, like Battle Creek, are represented by a single entry in 1894. By contrast, Pittsburgh has a publication with at least one article in the data in 36 different years. 

Without further standardization, it's hard to know whether the ranking of top cities is meaningful. The newspaper city NA problem is likely correctable. For missing values, another column in the dataset -- newspaper_name -- likely presents an opportunity to impute a newspaper's city. 

In almost all of the cases where newspaper_city is missing, the newspaper_name column contains the city name in parentheses after the publication title. These three examples are representative: 'Afro-American courier. (Yazoo City, Mis...)', 'Arizona sun (Phoenix, Ariz.)', 'Arizona tribune (Phoenix, Ariz.)', 'Cayton's weekly (Seattle, Wash.)', 'Chicago world. (Chicago, Ill.)'. String functions can be used to extract the city name and populate the newspaper city column. 

There are key unknowns that must be examined before using this sample to develop an understanding of Black newspaper articles. It's unclear whether the sample of articles here is temporally, topically or geographically representative of a complete corpus of Black newspaper articles.  

The dataset appears to be structured to enable content analysis on the text of each article.  The filepath and url columns appear to link to, respectively, the article text and the source article.  Those files could be loaded and the text extracted into a format suitable for standard content analysis methods.